---
title: "AM10 SG11 Group Project Analyses"
output: html_document
author: "Study Group 11: Alberto Lambert, Sijia Liu, Abhinav Bhardwaj, Bartek Makuch, Anna Plaschke, Feiyang Ni"
date: "2020-11-15"
output:
  html_document:
    theme: flatly
    highlight: zenburn
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: show
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width=8, 
  fig.height=5,
  fig.align = "left"
)
```

#Load Necessary Libraries

```{r, load_libraries, include = FALSE}
library(readr) #load dataset
library(tidyverse)
library(lubridate)
library(janitor)
library(skimr)
library(sf)
library(ggmap)
library(ggrepel)
library(gridExtra)
library(pander)
library(here)
library(leaflet)
library(tmap)
library(tmaptools)
library(hrbrthemes)
library(patchwork)
library(kableExtra)
library(scales)
library(here)
library(janitor)
library(broom)
library(tidyquant)
library(infer)
library(openintro)
library(knitr)
library(patchwork)
library(scales)
library(tidytext)
library(plotly)
library(corrplot)
library(leaflet)
library(rsample) 
library(ggfortify)
library(car)
library(huxtable)
library(Hmisc) #describe function
library(here)
library(vroom)
library(ggthemes)
library(cluster)
library(ggtext)
library(caret)

```

# Load Data

```{r loading_data, include = FALSE}
#loading first data set, incl. GDP
owid_covid_data <- read_csv(here::here("owid-covid-data.csv")) 
#cleaning the names
covid_data <- owid_covid_data %>% 
  clean_names()

#loading second data set: testing data
covid_testing_all_observations <- read_csv(here::here("covid-testing-all-observations.csv"))
#cleaning the names
covid_testing<- covid_testing_all_observations %>% 
  clean_names()

#loading third data set: number of hospitalized patients

hospital_data_original <- vroom(here::here("data.csv")) 
#cleaning the names
hospital_data <- hospital_data_original %>% 
  clean_names()
```

# Clean data

## Dataset 1: owid-covid-data

We start with the `covid_data`, which has 48 covid-related indicators of a country in a given date. From the `skim` below, we observe:

* 17 completely empty columns 
* Incompleteness in almost all columns

```{r dataset1 skim}
skim(covid_data)
```

```{r dataset1 remove duplicates and empties}
#check for duplicates
covid_data %>% get_dupes(date, iso_code) # no duplicates 
#Remove empty columns and rows
covid_data_clean <- remove_empty(covid_data, which = c("rows","cols")) # 17 columns and 0 rows removed
```

We start to clean each columns. We can see there are some rows with `location = International`which makes our `iso_code` NA. Since our analysis mainly focus on each countries, we will delete these rows. For the `continent` variable, we found that its NA come from`location = 'World'`, which is actually derivable from other rows, thus we will delete them.

```{r dataset1 clean iso_code and continent}
# NAs in iso_code
head(covid_data_clean[is.na(covid_data$iso_code),])
# delete them
covid_data_clean <- 
  covid_data_clean %>% filter(!is.na(iso_code))
# NAs in continent
head(covid_data_clean[is.na(covid_data_clean$continent),])
covid_data_clean <- covid_data_clean %>% 
  filter(!is.na(continent))
```

For the `gdp_per_capita` column, there are 28 countries/territories with missing values. And they are missing across all the days. We impute some of them, with data from world bank if it is available, and leave the rest in NAs

```{r dataset1 clean gdp_per_capita}
# check if the gdp_per_capita of countries are NA all the time 
covid_data_clean %>% 
  filter(is.na(gdp_per_capita)) %>% 
  group_by(location) %>% 
  summarise(all_na = all(is.na(gdp_per_capita)))

covid_data_clean[covid_data_clean$location =='Andorra','gdp_per_capita'] = 45887
covid_data_clean[covid_data_clean$location =='Cuba','gdp_per_capita'] = 6816
covid_data_clean[covid_data_clean$location =='Faeroe Islands','gdp_per_capita'] = 48530
covid_data_clean[covid_data_clean$location =='Greenland','gdp_per_capita'] = 49311
covid_data_clean[covid_data_clean$location =='Guam','gdp_per_capita'] = 31477
covid_data_clean[covid_data_clean$location =='Isle of Man','gdp_per_capita'] = 90190
covid_data_clean[covid_data_clean$location =='Liechtenstein','gdp_per_capita'] = 141200
covid_data_clean[covid_data_clean$location =='Monaco','gdp_per_capita'] = 196061
covid_data_clean[covid_data_clean$location =='Northern Mariana Islands','gdp_per_capita'] = 18577
covid_data_clean[covid_data_clean$location =='Turks and Caicos Islands','gdp_per_capita'] = 21028
covid_data_clean[covid_data_clean$location =='United States Virgin Islands','gdp_per_capita'] = 11069

skim(covid_data_clean)
```
##  Dataset 2: Hospital Data

```{r dataset2 skim}
skim(hospital_data)
```

We remove all rows with no indicator as we don't know what it would stand for in cases where `indicator` = NA, similarly we remove rows with NA in `value`

```{r dataset2 clean indicator and value}
hospital_data_clean <- hospital_data %>%
  filter(!is.na(indicator))
hospital_data_clean <- hospital_data_clean %>%
  filter(!is.na(value))
```

We then deselect some columns and pivot hospital data to a clean format

```{r clean data format}

hospital_data_clean <- hospital_data_clean %>% 
  pivot_wider(names_from=indicator, values_from=value) %>% 
  select(c(1,2,6,7)) %>% 
  clean_names() %>% 
  group_by(date, country) %>% 
  summarise(daily_hospital_occupancy = sum(daily_hospital_occupancy),
            daily_icu_occupancy = sum(daily_icu_occupancy))

hospital_data_clean %>% get_dupes(date, country) # no duplicates

```

## Dataset 3: COVID Testing Dataset

We explore the dataset...

```{r explore testing}
skim(covid_testing)

# we can drop a few variables such as source_url, source_label and note

covid_testing_process <- covid_testing %>% 
  select(-c(source_url, source_label, notes))
         
skim(covid_testing_process)

#show a couple of entries for specific country

covid_testing_process %>% 
  filter(iso_code == "POL") %>% 
  arrange(desc(date)) %>% 
  head(5)

```
As we can observe there are two metrics: people tested and samples tested. Having browsed the source of the dataset we are not able to conclude which data is correct, thus we are going to take daily average for specific dates.

```{r clean testing}

#average daily data per country and omit na's

covid_testing_process1 <-covid_testing_process %>% 
  group_by(date, iso_code) %>% 
  summarise(cumulative_tests = mean(cumulative_total, na.rm = TRUE),
            daily_tests = mean(daily_change_in_cumulative_total, na.rm = TRUE),
            cumulative_tests_per_thousand = mean(cumulative_total_per_thousand, na.rm = TRUE),
            x7_day_daily_test_change = mean(x7_day_smoothed_daily_change, na.rm = TRUE),
            x7_day_daily_test_change_per_thousand = mean(x7_day_smoothed_daily_change_per_thousand, na.rm = TRUE),
            short_term_tests_per_case = mean(short_term_tests_per_case, na.rm = TRUE),
            short_term_positive_rate = mean(short_term_positive_rate, na.rm = TRUE))

#check the results

covid_testing_process1 %>% 
  filter(iso_code == "USA") %>% 
  arrange(desc(date)) %>% 
  head(5)

```

## Merging the 3 Datasets into 1

```{r merge dataset}

#rename columns in hospital dataset in order to join datasets

hospital_data_clean <- hospital_data_clean %>% 
  rename(location = country)

#merging datasets

covid_total <- covid_data_clean %>% 
  left_join(hospital_data_clean, by= c("location", "date")) %>% 
  left_join(covid_testing_process1, by = c("date", "iso_code"))

#see what's inside

skim(covid_total)

```

We observe that there are still quite a few missing values, however we are going to focus on Europe in our analysis.

```{r select europe}
#filter for European countries

unique(covid_total$continent)

covid_europe <- covid_total %>% 
  filter(continent == "Europe") %>% 
  clean_names()

skim(covid_europe)
```
## Enriching dataset

We are going to enrich the dataset by defining the following variables:  

- day0_cases -> date when daily cases first hit 1000  
- day0_deaths -> date when daily deaths first hit 100  
- day0_cases_elapsed -> days elapsed from day_0_cases  
- day0_deaths_elapsed -> days elapsed from day_0_deaths  
- day0_difference -> days elapsed between daily cases first hit 1000 and daily deaths 100

```{r enriching the dataset}

##days elapsed since daily cases exceeded 1000 for the first time

#find the date zero
calc_day0 <- covid_europe %>% 
  filter(new_cases >= 1000) %>% 
  group_by(location) %>% 
  summarise(day0_cases = min(date))

#merge data with zero date
covid_europe_process <- covid_europe %>% 
  left_join(calc_day0, by = "location") %>% 

#add metric counting days from day 0

  mutate(day0_cases_elapsed = as.numeric(date - day0_cases)) %>% 
  
  #assign NA for negative days
  mutate(day0_cases_elapsed = case_when(
    day0_cases_elapsed < 0 ~ NA_real_,
    TRUE ~ day0_cases_elapsed
  ))

##days elapsed since daily deaths exceeded 100 for the first time

#find the date zero
calc_day0 <- covid_europe %>% 
  filter(new_deaths >= 100) %>% 
  group_by(location) %>% 
  summarise(day0_deaths = min(date))

#merge data with zero date
covid_europe_process2 <- covid_europe_process %>% 
  left_join(calc_day0, by = "location") %>% 

#add metric counting days from day 0

  mutate(day0_deaths_elapsed = as.numeric(date - day0_deaths)) %>% 
  
  #assign NA for negative days
  mutate(day0_deaths_elapsed = case_when(
    day0_deaths_elapsed < 0 ~ NA_real_,
    TRUE ~ day0_deaths_elapsed
  ))

##days elapsed between daily cases hit 1000 and daily deaths hit 100

#use threshold for daily deaths (some countries exceeded 1000 daily cases, but not 100 dail deaths)

calc_day0 <- covid_europe_process2 %>% 
  filter(new_deaths >= 100) %>% 
  group_by(location) %>% 
  summarise(day0_difference = min(as.numeric(day0_deaths - day0_cases))) #use min in order to get only 1 result per country

#merge all data

covid_europe_process3 <- covid_europe_process2 %>% 
  left_join(calc_day0, by = "location")

covid_europe <- covid_europe_process3

#clean temporary objects

rm(calc_day0,covid_europe_process, covid_europe_process2, covid_europe_process3)
```

Adjustments made in the following two plots:

1. We changed the date to most recent instead of Nov 24 2020 to avoid NAs, because some countries may not have yet reported their cases in that day
2. We changed the color scale, using  https://gka.github.io/palettes/#/9|s|ffffe0,920000|ffffe0,ff005e,93003a|1|1
3. We added some country names on the map. If we show all of them that would be too busy. We can adjust the names of the countries we want to show by changing the `country_list`

#Building Visualisations to Tell Our Story

We begin by setting the scene of the 'state of COVID=19' in Europe, before exploring the relationship between government measures and deaths/cases, and delving into 4 distinct country case studies to understand how specific public policy decisions have translated into COVID-19 outcomes. Finally, we build a predictive model to propose an optimal stringency level for a country facing a certain number of cases, of a certain size, and so on. We conclude that those countries who have successfully contained the virus have adopted a strategy of X, Y, and Z and experienced X, Y, Z outcomes, versus those who have exercised less coordinated and sustained approaches. 

## Plot 1: Mapping the most recent date (not neccesarily Nov 24 2020) new cases data in Europe

```{r new_cases_1124}
library(ggplot2)
library(sf)
library(rnaturalearth)
my_scale <- c('#feedde','#fdd0a2','#fdae6b','#fd8d3c','#f16913','#d94801','#8c2d04')
country_list = list('Sweden','Russia','Finland','Poland','Italy','Sweden', 'England')
# get the world map
worldmap <- ne_countries(scale = 'medium', type = 'map_units', continent = 'europe',
                         returnclass = 'sf')
# get the euro map
europe_cropped <- st_crop(worldmap, xmin = -20, xmax = 60,
                                    ymin = 30, ymax = 70)

# combine map with covid european data at the most recent date
covid_europe_lateset <- covid_europe %>% 
  group_by(location) %>% 
  mutate(my_rank = order(date, decreasing = TRUE)) %>% 
  filter(my_rank == 1)

euro_joined <- left_join(europe_cropped, covid_europe_lateset, by = c("admin" = "location"))

# mapping the covid data
euro_joined %>% 
  ggplot() +
  geom_sf(aes(fill = new_cases)) +
  geom_sf_text(data = euro_joined[euro_joined$subunit %in% country_list,], aes(label = admin),size = 4) +
  labs(title = "Change here",subtitle = 'Most Recent New Cases in Europe',
       x='',y='', caption = 'Source:https://owid.data') +
  scale_fill_gradientn(name = NULL,colours = my_scale,
                       labels = c(0,'5K','10K','15K','20K','25K')) +
  theme_fivethirtyeight() + theme(legend.position = "right", legend.direction = 'vertical')
```

## Plot 2: Mapping the most recent date (not neccesarily Nov 24 2020) total cases

```{r}
my_scale <- c('#fee5d9','#fcbba1','#fc9272','#fb6a4a','#ef3b2c','#cb181d','#99000d')
# mapping the covid data
euro_joined %>% 
  ggplot() +
  geom_sf(aes(fill = total_cases)) +
  geom_sf_text(data = euro_joined[euro_joined$subunit %in% country_list,], aes(label = admin),
               size = 4) +
  labs(title = 'changehere',subtitle = "Most Recent Total Cases in Europe", x='',y='',
       caption = 'Source:https://owid.data') +
  scale_fill_gradientn(name = NULL,colours = my_scale,
                       labels = c(0,'0.5M','1M','1.5M','2M')) +
  theme_fivethirtyeight() + theme(legend.position = 'right', legend.direction = 'vertical')
```

## General Analysis: Visualizing the trend of new cases after the first day new case number exceed 1000 in the top 5 countries with most cases in Europe

```{r top 5 sum new_cases countries}
top_5_country <- covid_europe %>% 
  group_by(location) %>% 
  summarise(sum = sum(new_cases)) %>% 
  arrange(desc(sum)) %>% 
  top_n(5) %>% 
  select(location) %>% 
  pull()
# Top 5 countries with most cases are France, Russia, Spain, United Kingdom and Italy
```

```{r redate from 1000 cases}
# rerank the date by setting the first date as the day first 1000 new cases diagnosed
covid_europe_1000 <- covid_europe %>% 
  filter(new_cases >= 1000) %>% 
  group_by(location) %>% 
  summarise(date_min = min(date)) %>% 
  mutate(start_min = as.Date(date_min))
covid_europe_1000

# join the start date and calculate the date diff to rank the days after first 1000 cases and create a new variable calls date_rnk
covid_europe_start <- left_join(covid_europe,covid_europe_1000,by = c("location" = "location"))
covid_europe_start <- covid_europe_start %>% 
  mutate(date_rnk = as.double(difftime(date,start_min,units = "days"))) %>% 
  filter(date_rnk >= 0)
```

```{r plot top 5 countries}
covid_europe_start_5 <- covid_europe_start %>% 
  filter(location %in% top_5_country)

#covid_europe_start_5 %>% 
  #ggplot(aes(x = date_rnk, y = new_cases_smoothed, color = location)) +
  #geom_line(size=0.5) +
  #labs(title = "The trend of new cases after the first day new case number exceed 1000 in the top 5 countries with most cases in Europe") +
  #theme_bw()
```

## Plot 3: Investigating the relationship between stringency measures and cases/deaths for prominent European countries
```{r}

#(1) New Cases Smoothed, Per Million

coeff=7.5

covid_europe %>% 
  group_by(location) %>% 
  filter(location %in% c("Denmark","Finland","France","Germany","Greece","Iceland","Italy","Netherlands","Norway","Poland","Slovakia","Slovenia","Spain","Sweden","Switzerland","United Kingdom")) %>% 
  
  ggplot(aes(x=date))+
  facet_wrap(~location)+
  geom_line(aes(y=new_cases_smoothed_per_million),colour="blue")+
  geom_line(aes(y=stringency_index*coeff), colour="red")+
  scale_y_continuous(name = "Smoothed New COVID-19 Cases Per Million",
    # Add a second axis and specify its features
    sec.axis = sec_axis( trans=~./coeff, name="Stringency Index")
  )+
  theme_fivethirtyeight()+
  theme(axis.title = element_text())+
  labs(title="Too Little Too Late? European Countries' Measures Suggest Overconfidence Before Major Resurgences of COVID-19", subtitle="Relationship between New Daily COVID-19 Cases and National Stringency Index",x="Date")+
  theme(legend.title=element_blank()) +
  theme(legend.position="top") 

# (2) New Deaths Smoothed, Per Million

covid_europe %>% 
  group_by(location) %>% 
  filter(location %in% c("Denmark","Finland","France","Germany","Greece","Iceland","Italy","Netherlands","Norway","Poland","Slovakia","Slovenia","Spain","Sweden","Switzerland","United Kingdom")) %>% 
  
  ggplot(aes(x=date))+
  facet_wrap(~location)+
  geom_line(aes(y=new_deaths_smoothed_per_million),colour="blue")+
  geom_line(aes(y=stringency_index/coeff), colour="red")+
  scale_y_continuous(name = "Smoothed New COVID-19 Deaths Per Million",
    # Add a second axis and specify its features
    sec.axis = sec_axis( trans=~.*coeff, name="Stringency Index")
  )+
  theme_fivethirtyeight()+
  theme(axis.title = element_text())+
  labs(title="Early Bird Catches the Worm | Countries with Strict, Rapid and Sustained Stringency Measures Culled COVID-19 Deaths", subtitle="Relationship between New Daily COVID-19 Deaths and National Stringency Index",x="Date")+
  theme(legend.title=element_blank()) +
  theme(legend.position="top") 


#(3) New Cases Smoothed, Per Million - For Days Since Surpassing 1000 Cases

coeff=10

covid_europe_start %>% 
  group_by(location) %>% 
  filter(location %in% c("Denmark","Finland","France","Germany","Greece","Iceland","Italy","Netherlands","Norway","Poland","Slovakia","Slovenia","Spain","Sweden","Switzerland","United Kingdom")) %>% 
  
  ggplot(aes(x=date_rnk))+
  facet_wrap(~location)+
  geom_line(aes(y=new_cases_smoothed_per_million),colour="blue")+
  geom_line(aes(y=stringency_index*coeff), colour="red")+
  scale_y_continuous(name = "Smoothed New COVID-19 Cases Per Million",
    # Add a second axis and specify its features
    sec.axis = sec_axis( trans=~./coeff, name="Stringency Index")
  )+
  theme_fivethirtyeight()+
  theme(axis.title = element_text())+
  labs(title="XXX", subtitle="Relationship between Smoothed New COVID-19 Cases Since Surpassing 1000 Cases and National Stringency Index",x="Days Since Reporting 1000 Cases")+
  theme(legend.title=element_blank()) +
  theme(legend.position="top") 

```

## Plot 4.1: (UK) Visualizing the trend of cases and tests alongside key policy events 

```{r UK CHART}
  
UK_1wave <- covid_europe %>% 
  # to create first wave, we filter the month
  filter(location == 'United Kingdom' & month(date)<9 & month(date)>1) %>% 
  ggplot() +
  # create tiles for the first lockdown
  geom_rect(aes(xmin = as.Date('2020-03-23'), xmax = as.Date('2020-05-13'), 
                ymin = -Inf, ymax=Inf), fill = 'grey90') +
  
  # first lockdown reference line
  geom_vline(xintercept = as.Date('2020-03-23'), color = 'darkgrey', linetype = 'dashed') +
  # first lockdown end reference line
  geom_vline(xintercept = as.Date('2020-05-13'), color = 'darkgrey', linetype = 'dashed') +
  
  # second lockdown reference line
  geom_vline(xintercept = as.Date('2020-11-05'), color = 'darkgrey', linetype = 'dashed') +
  
  # add annotate
  annotate(geom="text", x = as.Date('2020-04-17'),y = 18000, label='First\nLockdown', color = 'black') +
  
  # new cases left axis line
  geom_line(aes(x = date, y = new_cases_smoothed, group = 1,color = 'Cases')) +
  
  # tests right axis line
  # we devide y by 8 s.t it doesn't distort the graph
  geom_line( aes(x = date,y = x7_day_daily_test_change/8, colour = "Tests")) +
  
  # scale y axis, change right axis back to original values
  scale_y_continuous(name = 'Daily Cases',labels = scales::comma,sec.axis = sec_axis(~.*8, name = "Daily Tests",labels = scales::comma)) +
  
  theme_classic() +
  
  labs(subtitle = 'UK first wave') +
  scale_x_date(labels = date_format("%m-%Y"), date_breaks = "month") 

UK_2wave  <- covid_europe %>% 
  filter(location == 'United Kingdom' & month(date)>=9) %>% 
  ggplot() +
  # New restriction (Alert level introduced in UK)
  geom_vline(xintercept = as.Date('2020-10-14'), color = 'darkgrey', linetype = 'dashed') +
  annotate(geom = 'text', x = as.Date('2020-10-23'), y = 30000, label = 'New\nRestrictions', color = 'black') + 
  # Second lockdown
  geom_vline(xintercept = as.Date('2020-11-05'), color = 'red') +
  annotate(geom="text", x = as.Date('2020-11-13'),y = 30000, label='Second\nLockdown', color = 'black') +
  
  geom_line(aes(x = date, y = new_cases_smoothed, group = 1,color = 'Cases')) +
  geom_line( aes(x = date,y = x7_day_daily_test_change/8, colour = "Tests")) +
  scale_y_continuous(name = 'Daily Cases',labels = scales::comma,sec.axis = sec_axis(~.*8, name = "Daily Tests",labels = scales::comma)) +
  theme_classic() +
  
  labs(subtitle = 'UK second wave',x = '') +
  scale_x_date(labels = date_format("%m-%Y"), date_breaks = "month", 
               limits = c(as.Date('2020-09-01'),as.Date('2020-11-24'))) 
UK_1wave
UK_2wave
UK_1wave /UK_2wave + plot_annotation(
  title = 'Some countries had it under control',
  subtitle = 'And some not...', #adding a separate line for clarity
  caption = 'Source:https://owid.data')
```
## Plot 4.2: (Poland) Visualizing the trend of cases and tests alongside key policy events 

```{r POLAND CHART}

poland_1wave <- covid_europe %>% 
  filter(location == 'Poland' & month(date) < 9) %>% 
  ggplot(aes(x = date)) +
  
    #lockdown and first easing
  geom_rect(aes(xmin = as.Date('2020-03-25'), xmax = as.Date('2020-05-04'), 
                ymin = -Inf, ymax=Inf), fill = 'grey90') +
  geom_line( aes(y = new_cases_smoothed, colour = "Daily Cases")) +
  geom_line( aes(y = x7_day_daily_test_change/20, colour = "Daily Tests")) + #divide daily tests by 10 in order not to distort graph
  scale_y_continuous(labels = scales::comma ,sec.axis = sec_axis(~.*20, name = "Daily Tests", labels = scales::comma)) + #multiply secondary axis by 10 in order to show actual results
  labs(title = 'Poland') +
  scale_x_date(labels = date_format("%m-%Y"), date_breaks = "month") +

  
  #lockdown line
  geom_vline(xintercept = as.Date('2020-03-25'), color = 'darkgrey', linetype = 'dashed')+ 
  
  #ease of lockdown
  geom_vline(xintercept = as.Date('2020-05-04'), color = 'darkgrey', linetype = 'dashed')+ #add dashed line
  annotate(geom="text", x = as.Date('2020-04-15'),y = 1000,label='First\nLockdown') +
  
  theme_classic() +
  labs(y = "Daily Cases")

poland_2wave<- covid_europe %>% 
  filter(location == 'Poland' & month(date) > 8) %>% 
  ggplot(aes(x = date)) +

  geom_line( aes(y = new_cases_smoothed, colour = "Daily Cases")) +
  geom_line( aes(y = x7_day_daily_test_change/2, colour = "Daily Tests")) + #divide daily tests by 2 in order not to distort graph
  scale_y_continuous(labels = scales::comma ,sec.axis = sec_axis(~.*2, name = "Daily Tests", labels = scales::comma)) + #multiply secondary axis by 2 in order to show actual results
  labs(title = 'Poland') +
  scale_x_date(labels = date_format("%m-%Y"), date_breaks = "month") +

  
  #new restrictions
  geom_vline(xintercept = as.Date('2020-10-09'), color = 'darkgrey', linetype = 'dashed')+ #add dashed line
  annotate(geom="text", x = as.Date('2020-10-16'),y = 25000,label='New\nRestrictions', size = 3.5) +
  
  
  #2nd lockdown
  geom_vline(xintercept = as.Date('2020-10-24'), color = 'red')+ 
  annotate(geom="text", x = as.Date('2020-10-30'),y = 25000,label='Second\nLockdown', size = 3.5) +
  
  theme_classic() +
  labs(y = "Daily Cases")

#using patchwork to depict 2 maps

patchwork <- poland_1wave / poland_2wave

#adding title and caption
patchwork + plot_annotation(
  title = 'Some countries had it under control',
  subtitle = 'And some not...', #adding a separate line for clarity
  caption = 'Source:https://owid.data',)
```

## Plot 4.3: (Italy) Visualizing the trend of cases and tests alongside key policy events 


```{r ITALY_CHART, fig.width=15, fig.height=10}
IT_1wave <- covid_europe %>% 
  # to create first wave, we filter the month
  filter(location == 'Italy' & month(date)<9 & month(date)>1) %>% 
  ggplot() +
  # create tiles for the first lockdown Zona rossa milano
  geom_rect(aes(xmin = as.Date('2020-02-21'), xmax = as.Date('2020-03-08'), 
                ymin = -Inf, ymax=Inf), fill = 'grey90') +
  # create tiles for the first lockdown Zona rossa nord total
   geom_rect(aes(xmin = as.Date('2020-03-08'), xmax = as.Date('2020-04-03'), 
                ymin = -Inf, ymax=Inf), fill = 'grey70') +
  
  # first lockdown reference line: Zona Rossa Milano
  geom_vline(xintercept = as.Date('2020-02-21'), color = 'darkgrey', linetype = 'dashed') +
  # first lockdown end reference line Zona rossa nord total
  geom_vline(xintercept = as.Date('2020-03-08'), color = 'darkgrey', linetype = 'dashed') +
  # first lockdown end reference line
  geom_vline(xintercept = as.Date('2020-04-03'), color = 'darkgrey', linetype = 'dashed') +
  
  # add annotate
  annotate(geom="text", x = as.Date('2020-02-28'),y = 9000, label='Lodi &\nPadua', color = 'black') +
  annotate(geom="text", x = as.Date('2020-03-22'),y = 9000, label='North\nItaly', color = 'black')+
  annotate(geom="text", x = as.Date('2020-03-10'),y = 10000, label='Lockdown', face= "bold", color = 'black')+
  
  # new cases left axis line
  geom_line(aes(x = date, y = new_cases_smoothed, group = 1,color = 'Cases')) +
  
  # tests right axis line
  # we devide y by 8 s.t it doesn't distort the graph
  geom_line( aes(x = date,y = x7_day_daily_test_change/8, colour = "Tests")) +
  
  # scale y axis, change right axis back to original values
  scale_y_continuous(name = 'Daily Cases',labels = scales::comma,sec.axis = sec_axis(~.*8, name = "Daily Tests",labels = scales::comma)) +
  
  theme_classic() +
   theme(legend.position = "top") +
  
  labs(subtitle = 'Italy first wave', colour="") +
  scale_x_date(labels = date_format("%m-%Y"), date_breaks = "month") 


IT_1wave


IT_2wave  <- covid_europe %>% 
  filter(location == 'Italy' & month(date)>=9) %>% 
  ggplot() +
  # Second lockdown
  # create tiles for second lockdown Zona rossa/arancione/gialla
   geom_rect(aes(xmin = as.Date('2020-11-03'), xmax = as.Date('2020-11-20'), 
                ymin = -Inf, ymax=Inf), fill = 'grey70') +
    # second lockdown reference line
 
  geom_vline(xintercept = as.Date('2020-11-03'), color = 'darkgrey', linetype = 'dashed') +
  annotate(geom="text", x = as.Date('2020-11-12'),y = 40000, label='Second Lockdown\nfor selected regions', color = 'black') +
  
  
  geom_line(aes(x = date, y = new_cases_smoothed, group = 1,color = 'Cases')) +
  geom_line( aes(x = date,y = x7_day_daily_test_change/8, colour = "Tests")) +
  scale_y_continuous(name = 'Daily Cases',labels = scales::comma,sec.axis = sec_axis(~.*8, name = "Daily Tests",labels = scales::comma)) +
  theme_classic() +
  theme(legend.position = "none") +
  labs(subtitle = 'Italy second wave',x = '', colour="") +
  scale_x_date(labels = date_format("%m-%Y"), date_breaks = "month", 
               limits = c(as.Date('2020-09-01'),as.Date('2020-11-24'))) 

IT_2wave
IT_1wave /IT_2wave + plot_annotation(
  title = 'Some countries had it under control',
  subtitle = 'And some not...', #adding a separate line for clarity
  caption = 'Source:https://owid.data')
```

First Lockdown

- Initial local lockdown in ten municipalities of the province of Lodi in Lombardy and one in the province of Padua in Veneto from 21.02.2020.
-	People under complete lockdown: 50,000 people
-	Measures: 
    Very strict: People were only allowed to leave their homes to buy supplies such as food and medicine, no train service was provided, no movement allowed, police cars blocked roads into and out of the quarantined area and erected barriers
     
Expansion to northern provinces
-	Lockdown expanded to most of northern Italy from 8.03 – 3.04
-	People under lockdown: over sixteen million people
-	Measures: depending on region different
    - no traveling from, to or within the affected areas,
    - no funerals and cultural events
    - min. distance of 1 metre from one another in public locations
    - but allowed: trains and planes to and from the region for "proven work needs, emergencies, or health reasons
    - Restaurants and cafes were permitted to open, between 6:00 and 18:00, while many other public locations such as gyms, nightclubs, museums and swimming pools were closed altogether
    -	"smart working processes" to permit their employees to work from home
    - allowance to impose fines or up to three months' jail for people caught leaving or entering the affected zone without permission. 


Second Lockdown: @ALBERTO ##HOW IN DEBTH SHOULD I GO REGARDING RESTRICTIONS??? 
- Italy is divided into different areas with different restrictions
    - red area
          - Abruzzo, Campania, Provincia autonoma di Bolzano, Toscana, Valle d'Aosta
          - very strict: restaurants, bars closed, people can move max 1 km from home
    - orange area
          - Basilicata, Calabria, Emilia Romagna, Friuli Venezia Giulia, Lombardia, Marche, Piemonte, Puglia, Umbria
          - restaurants: take-away, restaurants for hotelguest open
    
    - yellow area
          - Lazio, Liguria, Molise, Provincia autonoma di Trento, Sardegna, Sicilia, Veneto
          - restaurant etc. open from 5 to 18, bars closed
          - people can move freely from 5am to 10 pm

## Plot 4.4: (Sweden) Visualizing the trend of cases and tests alongside key policy events 

```{r SWEDEN CHART}

sweden_1wave <- covid_europe %>% 
  filter((location == 'Sweden') & (month(date) < 9)) %>% 
  ggplot(aes(x = date)) +
  
    #lockdown and first easing
  #geom_rect(aes(xmin = as.Date('2020-03-16'), xmax = as.Date('2020-06-13'), 
  #              ymin = -Inf, ymax=Inf), fill = 'grey90') +
  geom_line( aes(y = new_cases_smoothed, colour = "Daily Cases")) +
  geom_line( aes(y = x7_day_daily_test_change/20, colour = "Daily Tests")) + #divide daily tests by 10 in order not to distort graph
  scale_y_continuous(labels = scales::comma ,sec.axis = sec_axis(~.*20, name = "Daily Tests", labels = scales::comma)) + #multiply secondary axis by 10 in order to show actual results
  labs(title = 'Sweden') +
  scale_x_date(labels = date_format("%m-%Y"), date_breaks = "month") +

  
  #lockdown line
  geom_vline(xintercept = as.Date('2020-03-16'), color = 'darkgrey', linetype = 'dashed')+ 
  annotate(geom="text", x = as.Date('2020-04-16'),y = 1000,label='Domestic Travel\nBan',size=3.5)+
  
  #ease of lockdown
  geom_vline(xintercept = as.Date('2020-06-13'), color = 'darkgrey', linetype = 'dashed')+ #add dashed line
 
  
  theme_classic() +
  labs(y = "Daily Cases")


sweden_2wave <- covid_europe %>% 
  filter(location == 'Sweden' & month(date) > 8 & year(date)> 2019) %>% 
  ggplot(aes(x = date)) +

  geom_line( aes(y = new_cases_smoothed, colour = "Daily Cases")) +
  geom_line( aes(y = x7_day_daily_test_change/2, colour = "Daily Tests")) + #divide daily tests by 2 in order not to distort graph
  scale_y_continuous(labels = scales::comma ,sec.axis = sec_axis(~.*2, name = "Daily Tests", labels = scales::comma)) + #multiply secondary axis by 2 in order to show actual results
  labs(title = 'Sweden') +
  scale_x_date(labels = date_format("%m-%Y"), date_breaks = "month") +

  
  #new restrictions
  geom_vline(xintercept = as.Date('2020-10-09'), color = 'darkgrey', linetype = 'dashed')+ #add dashed line
  annotate(geom="text", x = as.Date('2020-10-16'),y = 25000,label='\nNew\nRestrictions', size = 3.5) +
  
  
  #2nd lockdown
  #geom_vline(xintercept = as.Date('2020-10-24'), color = 'red')+ 
  #annotate(geom="text", x = as.Date('2020-10-30'),y = 25000,label='Second\nLockdown', size = 3.5) +
  
  theme_classic() +
  labs(y = "Daily Cases")

#using patchwork to depict 2 maps

patchwork <- sweden_1wave / sweden_2wave

#adding title and caption
patchwork + plot_annotation(
  title = 'Sweden with no Lockdowns yet',
  subtitle = 'And some not...', #adding a separate line for clarity
  caption = 'Source:https://owid.data',)
```

# Plot 5 - RUNNING A STATISTICAL MODEL EVALUATING THE RELATIONSHIP BETWEEN STRINGENCY AND CASES/DEATHS IN ORDER TO PROPOSE THE OPTIMAL STRINGENCY MEASURE FOR A GIVEN COUNTRY AT A GIVEN TIME (THE PRESENT DAY / THE MOST RECENT DAY)

```{r defining variables}

# create variables: days elapsed since day0
country_day0 <- 
  covid_europe %>% 
  group_by(location) %>% 
  mutate(my_rank = order(date, decreasing = FALSE)) %>% 
  filter(my_rank == 1) %>% 
  select(c(location,date))

covid_europe_lm <- covid_europe %>% 
  left_join(country_day0,'location') %>% 
  mutate(days_elapsed = as.numeric(date.x-date.y))

```

```{r}

#Define control variables
control <- trainControl (
    method="cv",
    number=5,
    verboseIter=TRUE) #by setting this to true the model will report its progress after each estimation

#we are going to train the model and report the results using k-fold cross validation
model1_lm<-train(
    total_cases_per_million~aged_65_older+
            stringency_index:days_elapsed+stringency_index +
           stringency_index:location
           , data = covid_europe_lm,
   method = "lm",
    trControl = control, na.action=na.exclude
   )

# summary of the results
summary(model1_lm)



```

```{r LR Model Variable Importance}

# we can check variable importance as well
importance <- caret::varImp(model1_lm, scale=TRUE)
plot(importance)

```

```{r}

covid_europe %>% 
  group_by(location) %>% 
  filter(location %in% c("Denmark","Finland","France","Germany","Greece","Iceland","Italy","Netherlands","Norway","Poland","Slovakia","Slovenia","Spain","Sweden","Switzerland","United Kingdom")) %>% 
  ggplot(aes(y=stringency_index, color=location))+geom_boxplot()

```

