---
title: "AM10 SG11 Group Project Analyses"
output: html_document
author: "Study Group 11: Alberto Lambert, Sijia Liu, Abhinav Bhardwaj, Bartek Makuch, Anna Plaschke, Feiyang Ni"
date: "2020-11-15"
output:
  html_document:
    theme: flatly
    highlight: zenburn
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: show
---
```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width=8, 
  fig.height=5,
  fig.align = "left"
)
```

```{r, load_libraries, include = FALSE}
library(readr) #load dataset
library(tidyverse)
library(lubridate)
library(janitor)
library(skimr)
library(sf)
library(ggmap)
library(ggrepel)
library(gridExtra)
library(pander)
library(here)
library(leaflet)
library(tmap)
library(tmaptools)
library(hrbrthemes)
library(patchwork)
library(kableExtra)
library(scales)
library(here)
library(janitor)
library(broom)
library(tidyquant)
library(infer)
library(openintro)
library(knitr)
library(patchwork)
library(scales)
library(tidytext)
library(plotly)
library(corrplot)
library(leaflet)
library(rsample) 
library(ggfortify)
library(car)
library(huxtable)
library(Hmisc) #describe function
library(here)
library(vroom)

```

# Load data

```{r loading_data, include = FALSE}
#loading first data set, incl. GDP
owid_covid_data <- read_csv(here::here("owid-covid-data.csv")) 
#cleaning the names
covid_data <- owid_covid_data %>% 
  clean_names()

#loading second data set: testing data
covid_testing_all_observations <- read_csv(here::here("covid-testing-all-observations.csv"))
#cleaning the names
covid_testing<- covid_testing_all_observations %>% 
  clean_names()

#loading third data set: number of hospitalized patients

hospital_data_original <- vroom(here::here("data.csv")) 
#cleaning the names
hospital_data <- hospital_data_original %>% 
  clean_names()
```

# Clean data
## dataset: owid-covid-data

We start with the `covid_data`, which has 48 covid-related indicators of a country in a given date. From the `skim` below, we observe:

* 17 completely empty columns 
* Incompleteness in almost all columns

```{r dataset1 skim}
skim(covid_data)
```

```{r dataset1 remove duplicates and empties}
#check for duplicates
covid_data %>% get_dupes(date, iso_code) # no duplicates 
#Remove empty columns and rows
covid_data_clean <- remove_empty(covid_data, which = c("rows","cols")) # 17 columns and 0 rows removed
```

We start to clean each columns. We can see there are some rows with `location = International`which makes our `iso_code` NA. Since our analysis mainly focus on each countries, we will delete these rows. For the `continent` variable, we found that its NA come from`location = 'World'`, which is actually derivable from other rows, thus we will delete them.

```{r dataset1 clean iso_code and continent}
# NAs in iso_code
head(covid_data_clean[is.na(covid_data$iso_code),])
# delete them
covid_data_clean <- 
  covid_data_clean %>% filter(!is.na(iso_code))
# NAs in continent
head(covid_data_clean[is.na(covid_data_clean$continent),])
covid_data_clean <- covid_data_clean %>% 
  filter(!is.na(continent))
```

For the `gdp_per_capita` column, there are 28 countries/territories with missing values. And they are missing across all the days. We impute some of them, with data from world bank if it is availble, and leave the rest in NAs

```{r dataset1 clean gdp_per_capita}
# check if the gdp_per_capita of countries are NA all the time 
covid_data_clean %>% 
  filter(is.na(gdp_per_capita)) %>% 
  group_by(location) %>% 
  summarise(all_na = all(is.na(gdp_per_capita)))

covid_data_clean[covid_data_clean$location =='Andorra','gdp_per_capita'] = 45887
covid_data_clean[covid_data_clean$location =='Cuba','gdp_per_capita'] = 6816
covid_data_clean[covid_data_clean$location =='Faeroe Islands','gdp_per_capita'] = 48530
covid_data_clean[covid_data_clean$location =='Greenland','gdp_per_capita'] = 49311
covid_data_clean[covid_data_clean$location =='Guam','gdp_per_capita'] = 31477
covid_data_clean[covid_data_clean$location =='Isle of Man','gdp_per_capita'] = 90190
covid_data_clean[covid_data_clean$location =='Liechtenstein','gdp_per_capita'] = 141200
covid_data_clean[covid_data_clean$location =='Monaco','gdp_per_capita'] = 196061
covid_data_clean[covid_data_clean$location =='Northern Mariana Islands','gdp_per_capita'] = 18577
covid_data_clean[covid_data_clean$location =='Turks and Caicos Islands','gdp_per_capita'] = 21028
covid_data_clean[covid_data_clean$location =='United States Virgin Islands','gdp_per_capita'] = 11069

skim(covid_data_clean)
```

##  dataset: hospital data

```{r dataset2 skim}
skim(hospital_data)
```

We remove all rows with no indicator as we don't know what it would stand for in cases where `indicator` = NA, similarly we remove rows with NA in `value`

```{r dataset2 clean indicator and value}
hospital_data_clean <- hospital_data %>%
  filter(!is.na(indicator))
hospital_data_clean <- hospital_data_clean %>%
  filter(!is.na(value))
```

We then deselect some columns and pivot hospital data to a clean format

```{r clean data format}

hospital_data_clean <- hospital_data_clean %>% 
  pivot_wider(names_from=indicator, values_from=value) %>% 
  select(c(1,2,6,7)) %>% 
  clean_names() %>% 
  group_by(date, country) %>% 
  summarise(daily_hospital_occupancy = sum(daily_hospital_occupancy),
            daily_icu_occupancy = sum(daily_icu_occupancy))

hospital_data_clean %>% get_dupes(date, country) # no duplicates

```

## Testing dataset

Let's explore covid testing dataset.
```{r explore testing}
skim(covid_testing)

# we can drop a few variables such as source_url, source_label and note

covid_testing_process <- covid_testing %>% 
  select(-c(source_url, source_label, notes))
         
skim(covid_testing_process)

#show a couple of entries for specific country

covid_testing_process %>% 
  filter(iso_code == "POL") %>% 
  arrange(desc(date)) %>% 
  head(5)



```
As we can observe there are two metrics: people tested and samples tested. Having browsed the source of the dataset we are not able to conclude which data is correct, thus we are going to take daily average for specific dates.


```{r clean testing}

#average daily data per country and omit na's

covid_testing_process1 <-covid_testing_process %>% 
  group_by(date, iso_code) %>% 
  summarise(cumulative_tests = mean(cumulative_total, na.rm = TRUE),
            daily_tests = mean(daily_change_in_cumulative_total, na.rm = TRUE),
            cumulative_tests_per_thousand = mean(cumulative_total_per_thousand, na.rm = TRUE),
            x7_day_daily_test_change = mean(x7_day_smoothed_daily_change, na.rm = TRUE),
            x7_day_daily_test_change_per_thousand = mean(x7_day_smoothed_daily_change_per_thousand, na.rm = TRUE),
            short_term_tests_per_case = mean(short_term_tests_per_case, na.rm = TRUE),
            short_term_positive_rate = mean(short_term_positive_rate, na.rm = TRUE))

#check the results

covid_testing_process1 %>% 
  filter(iso_code == "USA") %>% 
  arrange(desc(date)) %>% 
  head(5)




```


## Merging datasets


```{r merge dataset}

#rename columns in hospital dataset in order to join datasets

hospital_data_clean <- hospital_data_clean %>% 
  rename( location = country)

#merging datasets

covid_total <- covid_data_clean %>% 
  left_join(hospital_data_clean, by= c("location", "date")) %>% 
  left_join(covid_testing_process1, by = c("date", "iso_code"))

#see what's inside

skim(covid_total)

```

We observe that there are still quite a few missing values, however we are going to focus on Europe in our analysis.

```{r select europe}
#filter for European countries

unique(covid_total$continent)

covid_europe <- covid_total %>% 
  filter(continent == "Europe") %>% 
  clean_names()

skim(covid_europe)
```
## Enrich dataset

We are going to enrich the dataset by following variables:  
-day0_cases -> date when daily cases first hit 1000  
-day0_deaths -> date when daily deaths first hit 100  
-day0_cases_elapsed -> days elapsed from day_0_cases  
-day0_deaths_elapsed -> days elapsed from day_0_deaths  
-day0_difference -> days elapsed between daily cases first hit 1000 and daily deaths 100

```{r enrich dataset}

##days elapsed since daily cases exceeded 1000 for the first time

#find the date zero
calc_day0 <- covid_europe %>% 
  filter(new_cases >= 1000) %>% 
  group_by(location) %>% 
  summarise(day0_cases = min(date))

#merge data with zero date
covid_europe_process <- covid_europe %>% 
  left_join(calc_day0, by = "location") %>% 

#add metric counting days from day 0

  mutate(day0_cases_elapsed = as.numeric(date - day0_cases)) %>% 
  
  #assign NA for negative days
  mutate(day0_cases_elapsed = case_when(
    day0_cases_elapsed < 0 ~ NA_real_,
    TRUE ~ day0_cases_elapsed
  ))

##days elapsed since daily deaths exceeded 100 for the first time


#find the date zero
calc_day0 <- covid_europe %>% 
  filter(new_deaths >= 100) %>% 
  group_by(location) %>% 
  summarise(day0_deaths = min(date))

#merge data with zero date
covid_europe_process2 <- covid_europe_process %>% 
  left_join(calc_day0, by = "location") %>% 

#add metric counting days from day 0

  mutate(day0_deaths_elapsed = as.numeric(date - day0_deaths)) %>% 
  
  #assign NA for negative days
  mutate(day0_deaths_elapsed = case_when(
    day0_deaths_elapsed < 0 ~ NA_real_,
    TRUE ~ day0_deaths_elapsed
  ))

##days elapsed between daily cases hit 1000 and daily deaths hit 100

#use threshold for daily deaths (some countries exceeded 1000 daily cases, but not 100 dail deaths)

calc_day0 <- covid_europe_process2 %>% 
  filter(new_deaths >= 100) %>% 
  group_by(location) %>% 
  summarise(day0_difference = min(as.numeric(day0_deaths - day0_cases))) #use min in order to get only 1 result per country


#merge all data


covid_europe_process3 <- covid_europe_process2 %>% 
  left_join(calc_day0, by = "location")

covid_europe <- covid_europe_process3


#clean temporary objects

rm(calc_day0,covid_europe_process, covid_europe_process2, covid_europe_process3)





```






Adjustments made in the following two plots:
1. I changed the date to most recent instead of nov 24 2020 to avoid NAs, because some countries may have yet reported their cases in that day
2. Changed the color scale, using  https://gka.github.io/palettes/#/9|s|ffffe0,920000|ffffe0,ff005e,93003a|1|1
3. Added some country names on the map. if we show all of them that would be too busy. we can adjust the name of countries we want to show by changing the `country_list`

## Plot1: Mapping the most recent date (not neccesarily Nov 24 2020) new cases data in Europe

```{r new_cases_1124}
library(ggplot2)
library(sf)
library(rnaturalearth)
my_scale <- c('#feedde','#fdd0a2','#fdae6b','#fd8d3c','#f16913','#d94801','#8c2d04')
country_list = list('Germany','Russia','Finland','Poland','italy','Sweden')
# get the world map
worldmap <- ne_countries(scale = 'medium', type = 'map_units', continent = 'europe',
                         returnclass = 'sf')
# get the euro map
europe_cropped <- st_crop(worldmap, xmin = -20, xmax = 50,
                                    ymin = 30, ymax = 73)

# combine map with covid european data at the most recent date
covid_europe_lateset <- covid_europe %>% 
  group_by(location) %>% 
  mutate(my_rank = order(date, decreasing = TRUE)) %>% 
  filter(my_rank == 1)

euro_joined <- left_join(europe_cropped, covid_europe_lateset, by = c("name" = "location"))

# mapping the covid data
euro_joined %>% 
  ggplot() +
  geom_sf(aes(fill = new_cases)) +
  geom_sf_text(data = euro_joined[euro_joined$admin %in% country_list,], aes(label = admin),size = 3) +
  labs(title = "Most Recent New Cases in Europe",x='',y='') +
  scale_fill_gradientn(name = NULL,colours = my_scale,
                       labels = c(0,'5K','10K','15K','20K','25K')) +
  theme_bw() 
```
## Plot 1.2 Most recent total cases
```{r}
my_scale <- c('#fee5d9','#fcbba1','#fc9272','#fb6a4a','#ef3b2c','#cb181d','#99000d')
worldmap <- ne_countries(scale = 'medium', type = 'map_units', continent = 'europe',
                         returnclass = 'sf')

europe_cropped <- st_crop(worldmap, xmin = -20, xmax = 50,
                                    ymin = 30, ymax = 73)
covid_europe_lateset <- covid_europe %>% 
  group_by(location) %>% 
  mutate(my_rank = order(date, decreasing = TRUE)) %>% 
  filter(my_rank == 1)


euro_joined <- left_join(europe_cropped, covid_europe_lateset, by = c("name" = "location"))
# mapping the covid data
euro_joined %>% 
  ggplot() +
  geom_sf(aes(fill = total_cases)) +
  geom_sf_text(data = euro_joined[euro_joined$admin %in% country_list,], aes(label = admin),size = 3) +
  labs(title = "Most Recent Total Cases in Europe", x='',y='') +
  scale_fill_gradientn(name = NULL,colours = my_scale,
                       labels = c(0,'0.5M','1M','1.5M','2M')) +
  theme_bw()
```

## Plot2: Visualize the trend of new cases after the first day new case number exceed 1000 in the top 5 countries with most cases in Europe
```{r top 5 sum new_cases countries}
top_5_country <- covid_europe %>% 
  group_by(location) %>% 
  summarise(sum = sum(new_cases)) %>% 
  arrange(desc(sum)) %>% 
  top_n(5) %>% 
  select(location) %>% 
  pull()
# Top 5 countries with most cases are France, Russia, Spain, United Kingdom and Italy
```

```{r redate from 1000 cases}
# rerank the date by setting the first date as the day first 1000 new cases diagnosed
covid_europe_1000 <- covid_europe %>% 
  filter(new_cases >= 1000) %>% 
  group_by(location) %>% 
  summarise(date_min = min(date)) %>% 
  mutate(start_min = as.Date(date_min))
covid_europe_1000

# join the start date and calculate the date diff to rank the days after first 1000 cases and create a new variable calls date_rnk
covid_europe_start <- left_join(covid_europe,covid_europe_1000,by = c("location" = "location"))
covid_europe_start <- covid_europe_start %>% 
  mutate(date_rnk = as.double(difftime(date,start_min,units = "days"))) %>% 
  filter(date_rnk >= 0)
```

```{r plot top 5 countries}
covid_europe_start_5 <- covid_europe_start %>% 
  filter(location %in% top_5_country)

covid_europe_start_5 %>% 
  ggplot(aes(x = date_rnk, y = new_cases, color = location)) +
  geom_line(size=0.5) +
  labs(title = "The trend of new cases after the first day new case number exceed 1000 in the top 5 countries with most cases in Europe") +
  theme_bw()
```



